---
title: "main effects paper"
author: "Rachel Tao"
date: "2/21/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggmap)
library(ggcorrplot)
library(gplots)
library(GGally)
library(knitr)
library(patchwork)
library(psych)
library(factoextra)
library(fst)
library(splines)
library(rnoaa)
library(mgcv)
library(MuMIn)

knitr::opts_chunk$set(
	fig.asp = 0.6,
	fig.width = 6,
	out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

## 1. Load data

```{r}
# read in pcp/nmf scores

MI_sources <- read_csv("./data/MI_pcp_nmf.csv")

# get number of years to use for df for date
yr_num <- length(unique(year(MI_sources$date)))

# add variables for temp and rh lags and set data types
MI_sources <- MI_sources %>% 
    mutate(
    temp_lag1 = lag(temp, 1),
    temp_lag2 = lag(temp, 2),
    temp_lag3 = lag(temp, 3),
    temp_3day = (temp_lag1 + temp_lag2 + temp_lag3)/3,
    rh_lag1 = lag(rh, 1),
    rh_lag2 = lag(rh, 2),
    rh_lag3 = lag(rh, 3),
    rh_3day = (rh_lag1 + rh_lag2 + rh_lag3)/3,
    DayofWeek = as_factor(DayofWeek),
    date = as.numeric(date))

# create quasipoisson function
hacked.quasipoisson <- function() {
  MI_model <- quasipoisson(link = "log")
  MI_model$aic <- poisson(link = "log")$aic
  MI_model
}

# check data frame
summary(MI_sources)

# remove missing values
MI_sources %>% na.omit %>% nrow()

```

## 2. Run quasi poisson model

106: no sure what is happening here. Adding a comment would be supper helpful
```{r}
# all sources linear
MI_model <- gam(MI_events ~                
                source_1 +               
                source_2 +               
                source_3 +                
                source_4 +                 
                source_5 +                 
                source_6 +   
                source_7 +
                s(temp, k = 5, fx = TRUE, bs = 'cr') +  
                s(temp_3day, k = 5, fx = TRUE, bs = 'cr') +    
                s(rh, k = 5, fx = TRUE, bs = 'cr') +           
                s(rh_3day, k = 4, fx = TRUE, bs = 'cr') +     
                DayofWeek +                
                s(date, k = 4*yr_num + 1, fx = TRUE, bs = 'cr'),   
                family = "quasipoisson",   
                data = MI_sources)

summary(MI_model)


QAIC(update(MI_model, family = hacked.quasipoisson), chat = 1.3175) # use chat from summary above
# 6837.946


```

## 3. View model results - create Table S1 and Figure 2

```{r}

# compute IQR to get IQR increase
IQR <- MI_sources %>%
  pivot_longer(source_1:source_7,
               names_to = "term",
               values_to = "score") %>% 
  group_by(term) %>% 
  summarize(iqr = IQR(score, na.rm = TRUE))
  
# compute rate ratios, percent change, and relabel sources
# used for Table S1
results_iqr <- broom::tidy(MI_model, parametric = TRUE) %>% 
  inner_join(IQR) %>% 
  mutate(rr = round(exp(iqr*estimate), 6),
         rr_lci = round(exp(iqr*(estimate - 1.96*std.error)), 6),
         rr_uci = round(exp(iqr*(estimate + 1.96*std.error)), 6),
         pct_change = 100*(rr - 1),
         pct_lci = 100*(rr_lci - 1),
         pct_uci = 100*(rr_uci - 1),
         outliers = "full data",
          term = plyr::mapvalues(term,
            from = c("source_1", 
                     "source_2", 
                     "source_3", 
                     "source_4", 
                     "source_5", 
                     "source_6", 
                     "source_7"),
            to = c("Crustal Dust", 
                   "Salt", 
                   "Traffic", 
                   "Regional", 
                   "Cadmium", 
                   "Chromium", 
                   "Barium"))
          )

#write.csv(results_iqr, "./experimental_results/results_iqr.csv")

# Create forest plot - Figure 2
ForestPlot_full <- ggplot(data = results_iqr, 
  aes(
      x = term,  
      y = pct_change,      
      ymin = pct_lci,     
      ymax = pct_uci)) +  
  geom_pointrange() +   
  geom_errorbar() +      
  geom_hline(aes(yintercept = 0), lty = 2) +
  xlab("") +    
  ylab(expression("Percent change in MI admission rate (95% CI)")) +
  theme_classic() +
  theme(
    axis.text.x = element_text(
      face = "bold", color = "black", size = 16, angle = 315),
    axis.title = element_text(face = "bold", size = 16),
    legend.title = element_blank(),
    legend.text = element_text(face = "bold", size = 24),
    legend.position = "right",
    axis.text.y = element_text(
      face = "bold", color = "black", size = 18)
    )

ForestPlot_full # produces the plot in the plots panel

```

## 4. Nonlinearities

Line 231: This line gives an error. I think this may be extra code from copy/pasting because it doesn't seem to fit with the data or to be needed in the subsequent steps. 

For each source, we modeled non-linearly with all other sources linear to assess for nonlinearity. We show below sources where there was evidence of nonlinearity.

Source 2 nonlinear - Figure S9
```{r}
# run model
nonlinear_2 <- gam(MI_events ~       
                source_1 +
                s(source_2, fx = FALSE, bs = 'cr') +  
                source_3 +
                source_4 +
                source_5 +
                source_6 +
                source_7 +
                s(temp, k = 5, fx = TRUE, bs = 'cr') +  
                s(temp_3day, k = 5, fx = TRUE, bs = 'cr') +    
                s(rh, k = 5, fx = TRUE, bs = 'cr') +           
                s(rh_3day, k = 4, fx = TRUE, bs = 'cr') +     
                DayofWeek +                
                s(date, k = 4*yr_num + 1, fx = TRUE, bs = 'cr'),   
                family = "quasipoisson",   
                data = MI_sources)

summary(nonlinear_2)

QAIC(update(nonlinear_2, family = hacked.quasipoisson), chat = 1.3221)
# 6825.012

# Use predict to create estimates and the corresponding confidence intervals.
pred <- predict(nonlinear_2, newdata = MI_sources, se.fit = TRUE, type = "terms")

data_fit <- as.data.frame(pred$fit) %>%
  clean_names() %>%
  select(s_source_2) %>%
  rename(beta = s_source_2)

data_se <- as.data.frame(pred$se) %>%
  clean_names() %>%
  select(s_source_2) %>%
  rename(se = s_source_2)

data_full <- cbind(data_fit, data_se, MI_sources) %>%
  arrange(source_2) %>%
  mutate(lci = exp(beta - 1.96 * se),
         uci = exp(beta + 1.96 * se),
         rate_ratio = exp(beta)) %>%
  mutate(model = name,
         disease = disease)

# Plot - Figure S9
data_full %>%
  ggplot(aes(x = source_2)) +
  theme_classic(base_line_size = 0.6, base_size = 20) +
  geom_path(aes(y = rate_ratio), size = 1.2) +
  geom_ribbon(aes(ymin = lci, ymax = uci), alpha = 0.3) +
  geom_hline(yintercept = 1, linetype = 2, size = 0.5) +
  geom_rug() +
  xlab(expression("Salt score", size = 20)) +
  ylab("Rate Ratio") +
  coord_trans(y = "log2") +
  theme(plot.title  = element_text(vjust = -5, hjust = 0.035)) +
  scale_y_continuous(limits = c(0.67, 1.40), breaks = seq(.7, 1.40, by = 0.1)) +
  theme(plot.title  = element_text(vjust = -5, hjust = 0.035))

```

## 5. Source 5 nonlinear - Figure S8

Line 280-295: Maybe change the object names so the objects from the code chunk above are not overwriting or cause errors by getting mix-up with the above objects
Line 189-312: Was there a reason not to have a model including both of the nonlinear sources with a spline?

```{r}
# run model
nonlinear_5 <- gam(MI_events ~    
                source_1 +
                source_2 +
                source_3 +
                source_4 +
                s(source_5, fx = FALSE, bs = 'cr') +  
                source_6 +
                source_7 +
                s(temp, k = 5, fx = TRUE, bs = 'cr') +  
                s(temp_3day, k = 5, fx = TRUE, bs = 'cr') +    
                s(rh, k = 5, fx = TRUE, bs = 'cr') +           
                s(rh_3day, k = 4, fx = TRUE, bs = 'cr') +     
                DayofWeek +                
                s(date, k = 4*yr_num + 1, fx = TRUE, bs = 'cr'),   
                family = "quasipoisson",   
                data = MI_sources)

summary(nonlinear_5)

QAIC(update(nonlinear_5, family = hacked.quasipoisson), chat = 1.3234)
# 6819.38

# Use predict to create estimates and the corresponding confidence intervals.
pred <- predict(nonlinear_5, newdata = MI_sources, se.fit = TRUE, type = "terms")

data_fit <- as.data.frame(pred$fit) %>%
  clean_names() %>%
  select(s_source_5) %>%
  rename(beta = s_source_5)

data_se <- as.data.frame(pred$se) %>%
  clean_names() %>%
  select(s_source_5) %>%
  rename(se = s_source_5)

data_full <- cbind(data_fit, data_se, MI_sources) %>%
  arrange(source_5) %>%
  mutate(lci = exp(beta - 1.96 * se),
         uci = exp(beta + 1.96 * se),
         rate_ratio = exp(beta))

# Plot - Figure S8
data_full %>%
  ggplot(aes(x = source_5)) +
  theme_classic(base_line_size = 0.6, base_size = 20) +
  geom_path(aes(y = rate_ratio), size = 1.2) +
  geom_ribbon(aes(ymin = lci, ymax = uci), alpha = 0.3) +
  geom_hline(yintercept = 1, linetype = 2, size = 0.5) +
  geom_rug() +
  xlab(expression("Cadmium score", size = 20)) +
  ylab("Rate Ratio") +
  coord_trans(y = "log2") +
  theme(plot.title  = element_text(vjust = -5, hjust = 0.035)) +
  scale_y_continuous(limits = c(0.67, 1.40), breaks = seq(.7, 1.40, by = 0.1)) +
  theme(plot.title  = element_text(vjust = -5, hjust = 0.035))

```

## 6. Sensitivity analysis: outliers

Sensitivity analysis with removed outliers:
create dataset with outliers 3SD or higher removed 
```{r}
replaceOutlier <- function(x, cutoff = 3) {
  x[abs(x) > cutoff*sd(x, na.rm = T)] <- NA_real_
  x
}

MI_sources_outliers <- MI_sources %>%
  mutate(
    source_1 = replaceOutlier(source_1),
    source_2 = replaceOutlier(source_2),
    source_3 = replaceOutlier(source_3),
    source_4 = replaceOutlier(source_4),
    source_5 = replaceOutlier(source_5),
    source_6 = replaceOutlier(source_6),
    source_7 = replaceOutlier(source_7))

```


Run model with outliers removed - Figure S10

365: Note that you are calculating the IQR of the data that includes outlier rather than MI_sources_outlier
368-373: if you are using the IQR from the whole data then this code is redundant because you already estimated the IQR previously and have the IQR object in the environment

```{r}
linear_out <- gam(MI_events ~                
                source_1 +  
                source_2 +
                source_3 +
                source_4 +
                source_5 +
                source_6 +
                source_7 +
                s(temp, k = 5, fx = TRUE, bs = 'cr') +  
                s(temp_3day, k = 5, fx = TRUE, bs = 'cr') +    
                s(rh, k = 5, fx = TRUE, bs = 'cr') +           
                s(rh_3day, k = 4, fx = TRUE, bs = 'cr') +     
                DayofWeek +                
                s(date, k = 4*yr_num + 1, fx = TRUE, bs = 'cr'),   
                family = "quasipoisson",   
                data = MI_sources_outliers,
                maxit = 10000)

summary(linear_out)

QAIC(update(linear_out, family = hacked.quasipoisson), chat = 1.3586)


IQR <- MI_sources %>%
  pivot_longer(source_1:source_7,
               names_to = "term",
               values_to = "score") %>% 
  group_by(term) %>% 
  summarize(iqr = IQR(score, na.rm = TRUE))
  
results_iqr_out <- broom::tidy(linear_out, parametric = TRUE) %>% 
  inner_join(IQR) %>% 
  mutate(rr = round(exp(iqr*estimate), 6),
         rr_lci = round(exp(iqr*(estimate - 1.96*std.error)), 6),
         rr_uci = round(exp(iqr*(estimate + 1.96*std.error)), 6),
         pct_change = 100*(rr - 1),
         pct_lci = 100*(rr_lci - 1),
         pct_uci = 100*(rr_uci - 1),
         outliers = "outliers excluded",
         term = plyr::mapvalues(term,
           from = c("source_1", 
                    "source_2", 
                    "source_3", 
                    "source_4", 
                    "source_5", 
                    "source_6", 
                    "source_7"),
           to = c("Crustal", 
                  "Salt", 
                  "Traffic", 
                  "Regional", 
                  "Cadmium", 
                  "Chromium", 
                  "Barium")))

# create forest plot - Figure S10
ForestPlot_outliers <- ggplot(data = results_iqr_out, 
  aes(x = term,  
      y = pct_change,      
      ymin = pct_lci,     
      ymax = pct_uci)) +  
      geom_pointrange() +   
      geom_errorbar() +      
    theme_classic() +
    geom_hline(aes(yintercept = 0), lty = 2) +
    xlab("") +    
    ylab(expression("Percent change in MI admission rate (95% CI)")) +
    theme(
    axis.text.x = element_text(
      face = "bold", color = "black", size = 16, angle = 315),
    axis.title = element_text(face = "bold", size = 16),
    legend.title = element_blank(),
    legend.text = element_text(face = "bold", size = 24),
    legend.position = "right",
    axis.text.y = element_text(
      face = "bold", color = "black", size = 18)
    )

ForestPlot_outliers # produces the plot in the plots panel
```

