---
title: "paper figures"
author: "Rachel Tao"
date: "2/22/2022"
output: html_document
---
```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)
library(ggmap)
library(ggcorrplot)
library(gplots)
library(GGally)
library(knitr)
library(patchwork)
library(psych)
library(factoextra)
library(fst)
library(splines)
library(rnoaa)
library(mgcv)
library(MuMIn)

knitr::opts_chunk$set(
	fig.asp = 0.6,
	fig.width = 6,
	out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

Prep data
```{r include=FALSE}
# load pcp/nmf scores
pcp_nmf_scores <- read_csv("./data/pcp_nmf_scores_offset.csv")

# load MI data (this is not actually in the repo)
MI <- read_fst("./data/z_NYC_daily_mi.fst") %>% 
  as_tibble() %>% 
  select(CaseDate, countMI.AdmitDate) %>% 
  # choose admit date instead of estimated time of event
  # to retain admissions with an admit date but missing hour of admission
  rename(
    date = CaseDate,
    MI_events = countMI.AdmitDate
  ) %>% 
  mutate(
    date = as.Date(date)
  ) %>% 
  filter(date > as.Date("2006-12-31"))

# load weather data
weather_to_2015 <- read_csv("./data/nyc_dailyweathervar.csv") %>% 
  as_tibble() %>% 
  mutate(
    day = round(day, digits = 0),
    day = ymd(day)) %>% 
  rename(date = day) %>% 
  filter(date > as.Date("2006-12-31")) 

# take averages to get daily values for rh and temp
weather_to_2015 <- weather_to_2015 %>% 
  group_by(date) %>% 
  summarize(rh = mean(rh), temp = mean(temp))

# join datasets together
MI_sources <- MI %>% 
  left_join(weather_to_2015, by = "date") %>% 
  left_join(pcp_nmf_scores, by = "date") %>% 
  mutate(population = 8175133,
         DayofWeek = as.character(wday(date, label = TRUE))
         ) %>% 
  rename(
    source_1 = V1,
    source_2 = V2,
    source_3 = V3,
    source_4 = V4,
    source_5 = V5,
    source_6 = V6,
    source_7 = V7
  )

# relevel days of the week
MI_sources$DayofWeek <- factor(MI_sources$DayofWeek,
                        level = c("Sun", 
                                  "Mon", 
                                  "Tue", 
                                  "Wed", 
                                  "Thu", 
                                  "Fri", 
                                  "Sat"))
# export dataset
# write_csv(MI_sources, "./data/MI_pcp_nmf.csv")

summary(MI_sources)
```

Table 1: summary statistics
```{r}
constituents <- read_csv("./data/nyc_raw.csv")

ng <- function(x, na.rm = TRUE) (x*1000)

summary <- left_join(MI_sources, constituents, by = "date") %>% 
  select(MI_events, rh, temp, pm25:zinc) %>% 
  mutate_at(5:30, ng, na.rm = TRUE)

table_1 <- summary %>% 
  pivot_longer(MI_events:zinc,
               names_to = "category",
               values_to = "value") %>% 
  group_by(category) %>% 
  summarize(
    min = min(value, na.rm = TRUE),
    Q1 = quantile(value, probs = 0.25, na.rm = TRUE),
    median = median(value, na.rm = TRUE),
    Q3 = quantile(value, probs = 0.75, na.rm = TRUE),
    max = max(value, na.rm = TRUE),
    mean = mean(value, na.rm = TRUE),
    sd = sd(value, na.rm = TRUE)
  ) %>% 
  mutate(
    category = as_factor(category),
    category = fct_recode(category,
                           "Aluminum (Al)" = "aluminum",
                           "Ammonium (NH4)" = "ammonium ion",
                           "Arsenic (As)" = "arsenic",
                           "Barium (Ba)" = "barium",
                           "Bromine (Br)" = "bromine",
                           "Cadmium (Cd)" = "cadmium",
                           "Calcium (Ca)" = "calcium",
                           "Chlorine (Cl)" = "chlorine",
                           "Chromium (Cr)" = "chromium",
                           "Copper (Cu)" = "copper",
                           "Elemental Carbon (EC)" = "elemental carbon",
                           "Iron (Fe)" = "iron",
                           "Lead (Pb)" = "lead",
                           "Magnesium (Mg)" = "magnesium",
                           "Manganese (Mn)" = "manganese",
                           "Nickel (Ni)" = "nickel",
                           "Organic Carbon (OC)" = "organic carbon",
                           "Potassium (K)" = "potassium ion",
                           "Selenium (Se)" = "selenium",
                           "Silicon (Si)" = "silicon",
                           "Sodium (Na)" = "sodium",
                           "Sulfur (S)" = "sulfur",
                           "Titanium (Ti)" = "titanium",
                           "Nitrate (NO3)" = "total nitrate",
                           "Vanadium (V)" = "vanadium",
                           "Zinc (Zn)" = "zinc",
                           "Total PM_[2.5]" = "pm25",
                           "Relative Humidity (%)" = "rh",
                           "Temperature (°C)" = "temp",
                           "MI Count" = "MI_events"
                           )) %>% 
      rename("Minimum" = "min",
             "25th%ile" = "Q1",
             "Median" = "median",
             "75th%ile" = "Q3",
             "Maximum" = "max",
             "Mean" = "mean")

#write_csv(table_1, "./table_1.csv")

```

Figure 1 is in root_ncnvx_pcp_paper
Figure 2 is in main_effects_paper
Table S1 is in main_effects_paper

Seasonal trends:
Table S2
Figures S2, S6
```{r include=FALSE}
MI_sources_long <- MI_sources %>% 
  pivot_longer(source_1:source_7,
               names_to = "source",
               names_prefix = "source_",
               values_to = "score") %>% 
  group_by(source) %>% 
  mutate(
    source = as_factor(source),
    source = fct_recode(source,
      "Crustal Dust" = "1", 
      "Salt" = "2",
      "Traffic" = "3",
      "Regional" = "4",
      "Cadmium" = "5",
      "Chromium" = "6",
      "Barium" = "7"
      ))

yr_num <- length(unique(year(MI_sources$date)))

# plot  - Figure S6
seasonal_trends <- ggplot(MI_sources_long, aes(date, score)) +
  geom_smooth(method = lm, formula = y ~ ns(x, df= 4*yr_num), color = "black") +
  facet_wrap(~source, scales = "free", ncol = 1) + 
  xlab("") +
  ylab("Factor Score") +
  theme_classic() +
  theme(
    axis.title.y = element_text(size = 18),
    strip.text = element_text(size = 14),
    axis.text.x = element_text(size = 13),
    axis.text.y = element_text(size = 11)
  ) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")

# plot MI trends - Figure S2
mi <- ggplot(MI_sources, aes(date, MI_events)) + 
  geom_smooth(method = lm, formula = y ~ ns(x, df= 4*yr_num), color = "black") +
  ylab("Number of MI Admissions") +
  xlab("") +
  theme_classic() +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 20)
  ) +
  scale_x_date(date_labels = "%Y", date_breaks = "1 year")

# extract season from dates
season <- MI_sources %>% mutate(season = metR::season(date))

# anova summary - repeat for each source
# Table S2
aov(source_1 ~ season, season) %>% summary()

# anova means - repeat for each source
# Table S2
aov_object <- aov(source_1 ~ season, season)

model.tables(aov_object, type = "means")

```


Day of week patterns:
Table S3, Figure S5
```{r include=FALSE}
# mean source scores for each day of the week
week <- MI_sources %>% 
  pivot_longer(source_1:source_7,
               names_to = "source",
               values_to = "score") %>% 
  group_by(source, DayofWeek) %>% 
  summarize(mean_score = mean(score, na.rm = TRUE)) %>% 
  mutate(
    source = as_factor(source),
    source = fct_recode(source,
      "Crustal Dust" = "source_1",   # *
      "Salt" = "source_2",           #
      "Traffic" = "source_3",        # *
      "Regional" = "source_4",       # *
      "Cadmium" = "source_5",        #
      "Chromium" = "source_6",       #
      "Barium" = "source_7"          #
      )
  )

# plot weekday trends - Figure S5
ggplot(week, aes(DayofWeek, mean_score)) +
  geom_bar(stat = "identity") +
  xlab("") +
  facet_wrap(~source, scales = "free", ncol =1) +
  ylab("Mean Factor Score") +
  ylim(0, 3) +
  theme_classic() +
  theme(
    axis.title.y = element_text(size = 18),
    axis.text.x.bottom = element_text(size = 13, color = "black"),
    strip.text = element_text(size = 14),
    axis.text.y.left = element_text(size = 12)
  )

# tag weekend vs. weekdays
week_day <- MI_sources %>% 
  mutate(week_end = if_else(DayofWeek %in% 
                              c('Mon', 'Tue', 'Wed', 'Thu', 'Fri'), 
                            'weekday', 'weekend'),
         week_end = as_factor(week_end))

# t-test for weekday vs. weekend means - Table S3
# repeat for each source
t.test(week_day$source_1~week_day$week_end)

# find mean and SD for weekend/weekday
# repeat for each source
week_day %>% group_by(week_end) %>% summarize(m = mean(source_1, na.rm = TRUE),
                                              sd = sd(source_1, na.rm = TRUE))

```

Figure S1: Map of monitor locations
```{r map, echo=FALSE, message=FALSE}
monitor_locations <- 
  read_csv("./data/monitor_sites.csv") %>% 
  mutate(site_number = as.character(site_number))

library(tigris)
library(sp)
library(rgdal)
library(sf)
library(leaflet)

lookup_code("New York", "New York")
lookup_code("New York", "Kings")
lookup_code("New York", "Queens")
lookup_code("New York", "Bronx")
lookup_code("New York", "Richmond")

nyc_counties = st_read("data/cb_2018_us_county_500k/cb_2018_us_county_500k.shp") %>%
  filter(STATEFP == "36",
         COUNTYFP %in% c('061','047','081','005','085')
         )

st_crs(monitor_locations)
st_crs(nyc_counties) # different

monitor_locations = merge(nyc_counties, monitor_locations)

st_crs(nyc_counties)

names(nyc_counties)

wgs84 <- CRS("+proj=longlat +datum=WGS84")

monitor_locations <- st_transform(monitor_locations, CRS("+proj=longlat +datum=WGS84"))

map <- leaflet(monitor_locations) %>% 
  addProviderTiles("CartoDB.PositronNoLabels") %>% 
  addCircleMarkers(lng = ~longitude, lat = ~latitude, label = ~site_number)

map

```

Figure S2 is above
Figure S3 is in root_ncnvx_pcp_paper

Figure S4: Pie chart
```{r}
# take the matrix norm of each source
pcp_nmf_source_1 <- pcp_nmf_scores %>% 
  select(V1) %>% 
  as.matrix() %>% 
  norm()

pcp_nmf_source_2 <- pcp_nmf_scores %>% 
  select(V2) %>% 
  as.matrix() %>% 
  norm()

pcp_nmf_source_3 <- pcp_nmf_scores %>% 
  select(V3) %>% 
  as.matrix() %>% 
  norm()

pcp_nmf_source_4 <- pcp_nmf_scores %>% 
  select(V4) %>% 
  as.matrix() %>% 
  norm()

pcp_nmf_source_5 <- pcp_nmf_scores %>% 
  select(V5) %>% 
  as.matrix() %>% 
  norm()

pcp_nmf_source_6 <- pcp_nmf_scores %>% 
  select(V6) %>% 
  as.matrix() %>% 
  norm()

pcp_nmf_source_7 <- pcp_nmf_scores %>% 
  select(V7) %>% 
  as.matrix() %>% 
  norm()

# add norms together to get total
tot = pcp_nmf_source_1 + pcp_nmf_source_2 + pcp_nmf_source_3 + pcp_nmf_source_4 + pcp_nmf_source_5 + pcp_nmf_source_6 + pcp_nmf_source_7

# calculate percentages
vals <- c((pcp_nmf_source_3/tot)*100, (pcp_nmf_source_4/tot)*100, (pcp_nmf_source_1/tot)*100, (pcp_nmf_source_7/tot)*100, (pcp_nmf_source_2/tot)*100, (pcp_nmf_source_5/tot)*100, (pcp_nmf_source_6/tot)*100)

#create vector of source names
val_names <- c("Traffic", "Regional", "Crustal", "Salt", "Cadmium", "Chromium", "Barium")

# label percentages vector with source names
names(vals) <- val_names

# make into tibble
values <- cbind(vals, val_names) %>% 
  as_tibble() %>% 
  mutate(vals = as.double(vals),
         vals = round(vals, digits = 0),
         val_names = fct_reorder(val_names, vals)
         )

# pie chart - Figure S4
values %>% 
  as_tibble() %>% 
  mutate(vals = as.double(vals)) %>% 
  ggplot(aes(x = "", y = vals, fill = val_names)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    theme_void() +
    theme(
      legend.title = element_blank(),
      legend.key.size = unit(1, 'cm'),
      legend.text = element_text(size = 14)
      ) +
    geom_text(aes(label = vals),
              position = position_stack(vjust = 0.5),
              size = 6)

```

Figure S5 and Figure S6 are above

Figure S7: Source correlation matrix
```{r}
# correlation matrix
pcp_nmf_scores %>% 
  rename(
    Crustal = V1,
    Salt = V2,
    Traffic = V3,
    Regional = V4,
    Cadmium = V5,
    Chromium = V6,
    Barium = V7
  ) %>% 
ggcorr(method = c("pairwise.complete.obs", "pearson"), limits = FALSE,
       label = TRUE, label_size = 5,
       hjust = 1, size = 6, color = "grey50", layout.exp = 4)

```

Figures S8, S9, and S10 are in main_effects_paper
Figure S11 is in root_ncnvx_pcp_paper